Proof by reflection for equalities in algebraic structures

### 2014-06-26 newPaper/paper.tex
paper

We will define a hierarchy of typeclasses. All our tactics (the group solver, the monoid solver...) will be able to work on any type, being given that an instance of the corresponding typeclass is provided.

@@

All our tactics will require to have a way of testing the equality between elements of the set, that is to say, a way to test equality between constants. For this reason, we define a notion of "set", which only requires this equality. All the algebraic structures above will extend this typeclass :

@@

Of course, there will be no tactic associated to $Set$, since we have no operations and not properties associated to this structure. Therefore, the equality in a $Set$ is just the syntactical equality, and can be proven with $refl$.

The first structure, almost trivial, is the magma. A magma has just a $Plus$ operation which let us computing constants, and no axioms about this operation. Thus, as we will see later, the normalization will be simply a computation between constants inside parenthesis.

@@

A bit more interesting is the Semi-Group typeclass. A semi-Group is a Magma (ie, it still has a $plus$ operation), but it also has the property of associativity for this operation.

@@

A Group is a Monoid, with two new operations. The binary $Minus$, and the unary $Neg$. One axiom says that the $Minus$ can always be simplified with the $Neg$ and the $Plus$. That means that $Minus$ is not a "primitive" operation of a Group.
The second axiom, which is the most important one for a Group, is the fact that any $c1$ admits $Neg c1$ as an inverse, where being an inverse if define just under.

@@

Note that the properties required by these typeclass for a specific type $T$ will either be proven if the operations on $T$ are real computable definitions, or axiomatised if the user is working in an axiomatised theory where the operations ($Plus$, $Neg$) are just axioms.

@@

We will have one datatype for reflected terms in each algebraic structure. \\
Blablabla

@@

For the moment, the definition of $Variable$ is just the following:

@@

And $get\_eq\_magma$ is just a function which returns the equality test of the $Set$ level.

>>>>

We will construct a hierarchy of typeclasses. All our tactics (the group solver, the monoid solver...) will be able to work on any type, being given that an instance of the corresponding typeclass is provided.

@@

All our tactics will require to have a way of testing the equality between elements of the underlying set, that is to say, a way to test equality between constants. For this reason, we define a notion of "set", which only requires this equality. All the algebraic structures above will extend this typeclass :

@@

Of course, there will be no tactic associated to $Set$, since we have no operations and not properties associated to this structure. Therefore, the equality in a $Set$ is just the syntactical equality, and can simply be proven with $refl$.

The first real structure, almost trivial, is the magma. A magma has just a $Plus$ operation which let us computing constants, and has no axioms about this operation. Thus, as we will see later, the normalization will be simply a computation between constants inside parenthesis.

@@

A bit more interesting is the Semi-Group typeclass. A semi-Group is a Magma (ie, it still has a $plus$ operation), but moreover it has the property of associativity for this operation.

@@

A Group is a Monoid, with two new operations. The binary operation $Minus$, and the unary operation $Neg$. One axiom says that the $Minus$ can always be simplified with the $Neg$ and the $Plus$. That means that $Minus$ is not a "primitive" operation of a Group, since we can always rewrite $(a-b)$ into $(a\ +\ -b)$.
The second axiom, which is the most important one for a Group, is the fact that any $c1$ admits $Neg\ c1$ as an inverse, where being an inverse if the following property :

@@

Note that the properties required by these typeclass for a specific type $T$ will either be proven if the operations on $T$ are real computable definitions, or axiomatised if the user is working on an axiomatised theory where the operations ($Plus$, $Neg$) are just axioms.

@@

We will have one datatype for reflected terms in each algebraic structure. \\
Each of these datatype is parametrised on a type $c$ : the real type on which we want to prove equalities. It is also indexed over an instance of the corresponding typeclass for $c$ (we usally call it $p$, because it looks like a proof saying that the structure $c$ has the desired properties), and indexed over a context (a vector $g$ of $n$ elements of type $c$), and also indexed over a value of type $c$, which is precisely the encoded value.
A magma is only equiped with one operation $Plus$. Thus, we only have three concepts to express in order to reflect terms in a Magma : constantes, variables, and addition.

@@

When we encode a constant $c1$ in a context $\Gamma$, we use the constructor $ConstMa$ to produce a term of type $ExprMa\ p\ \Gamma\ c1$ : the index representing the concrete value is simply the constant $c1$.
If e1 is an expression of type $ExprMa\ p\ \Gamma\ c1$ (ie, a term encoding the value $c1$), and $e2$ is an expression of type $ExprMa\ p\ \Gamma\ c2$ (a term encoding the value $c2$), then the term $PlusMa\ e1\ e2$ will have the type $ExprMa\ p\ \Gamma\ (Plus\ c1\ c2)$, ie, this term will encode the value $(Plus\ c1\ c2)$, where $Plus$ is the operation defined in the courant instance $p$.

We've used $get\_eq\_magma$ in the constructor $VarMa$, and this is just a function which returns the equality test of the $Set$ level.
For the moment, the definition of $Variable$ is just the following:

@@

This type will be extended with some other constructors later on.
Because there is no more operations in a SemiGroup or in a Monoid, the datatypes for reflected terms in these two structures will have exactly the same shape as the one for Magma that we've given above.
However, the one for $Group$ will introduce two new constructors for the $Neg$ and $Minus$ operations :

\begin{code}[caption=Reflected terms in a Group, captionpos=b, label=lst1:haskell2]
data ExprG :  Group c -> (Vect n c) -> c -> 
              Type where
    ConstG : (p : dataTypes.Group c) 
       -> (g:Vect n c)
       -> (c1:c) -> ExprG p g c1
    PlusG : {p : Group c} -> {g:Vect n c} 
       -> {c1:c} -> {c2:c} 
       -> ExprG p g c1 
       -> ExprG p g c2 
       -> ExprG p g (Plus c1 c2)
    MinusG : {p : Group c} -> {g:Vect n c} 
       -> {c1:c} -> {c2:c} 
       -> ExprG p g c1 
       -> ExprG p g c2 
       -> ExprG p g (Minus c1 c2)
    NegG : {p : Group c} -> {g:Vect n c} 
       -> {c1:c} -> ExprG p g c1 
       -> ExprG p g (Neg c1)
    VarG : (p : Group c) -> {g:Vect n c} 
       -> {c1:c}
       -> Variable (get_eq_group p) Neg g c1 
       -> ExprG p g c1
\end{code}
